Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-04 manuscript/manuscript.tex
Small cosmetic changes to R code. Review of last part of Manuscript

Collectively these six City agencies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the `big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

\begin{figure}[tbp]
    \centering
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs Created by Hour of the Day}
        \label{fig:busiestcreated}
    \end{subfigure}
    \par\medskip
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs Closed by Hour of the Day}
        \label{fig:busiestclosed}
    \end{subfigure}
    \caption{Comparison of SRs Created and Closed by Hour of the Day}
    \label{fig:stacked}
\end{figure}

@@

\dt{STOPPED EDITING HERE}

\textbf{agency \& agency\_name:} The \texttt{agency} field contains 
abbreviations of City agencies (e.g., NYPD, DOT), while 
\texttt{agency\_name} contains the full names. Including both seems 
redundant, especially given the space taken up by full names in the 
CSV file.

\jy{In my Chance article, we found an agency
  `DEPARTMENT OF CONSUMER AND WORKER PROTECTION', which occurred 361
  times in May 2023. It was the only one that does not look like an
  acronym. Did this occur in the two-year data?
  I wonder if this has been fixed lately.}

\jy{A related question is whether the data quality has improved over
  the years}

\texttt{landmark} \& \texttt{street\_name:} While the \texttt{landmark} field is 
intended for noteworthy locations (e.g., parks, airports), most entries 
are actually street names, with a 62\% match to \texttt{street\_name}. 
Even non-matches (excluding blanks) appear to be similar, differing 
only in spelling or formatting (e.g., ``NINTH AVE'' vs. ``9 AVE''). This 
indicates a possible misuse of the \texttt{landmark} field as well as
a high level of redundancy.

\texttt{cross\_street\_1} \& \texttt{intersection\_street\_1} and \texttt{cross\_street\_2} 
\& \texttt{intersection\_street\_2}: These two sets of street pairs are used to 
help identify the incident address. We found 88\% of 
these pairs to be duplicates. This duplication  
raises the question of which field-pair should be trusted. It is possible 
that the two street pairs originated from different agencies that used 
different variable names to represent the same information, but no 
documentation exists to clarify this relationship. Without question
these two street-pairs are highly redundant.

@@

Table~\ref{tab:summary1} presents a summary of the match and 
near-match results for \texttt{cross\_street\_1} and 
\texttt{intersection\_street\_1}. As the table shows, 88.15\% of the 
entries are exact matches, while 25.4\% are blank but matching.

@@

By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 39.5\%, which for this large dataset equates to 
a reduction of 1.4 Gb, which is significant. A smaller file size means faster downloads, 
less storage impact, and simpler data analysis efforts.
Below is a list of duplicate and near-duplicate fields.

\begin{itemize}
    \item \textbf{agency\_name}: Each row contains the \texttt{agency} field, 
    which uses abbreviations that are clear and well understood.

    \item \textbf{park\_borough}: A 100\% match with the \texttt{borough} field, 
    leading to no data loss.

    \item \textbf{location}: A straight concatenation of latitude and longitude 
    fields, offering no additional value beyond duplication.

    \item \textbf{cross\_street\_1 \& 2 and intersection\_street\_1 \& 2}: 
    An 88\% match between these pairs. We recommend deleting the 
    \texttt{intersection\_street\_1 \& 2} fields while acknowledging the
    concomitant data loss.

@@

Additionally, certain fields may not be useful due to their sparse population. 
For example, the \texttt{taxi\_company\_borough} field is 99.94\% blank, 
\texttt{road\_ramp} is 99.8\% blank, \texttt{vehicle\_type} is 99.7\% blank, 
and \texttt{due\_date} is 99.6\% blank. Several other fields, such as 
\texttt{bridge\_highway\_direction}, \texttt{bridge\_highway\_name}, 
\texttt{bridge\_highway\_segment}, and \texttt{taxi\_pick\_up\_location}, 
are over 99\% blank, with most used exclusively by the TLC.

Further reduction is possible by coding categorical 
variables, such as \texttt{complaint\_type}, through methods like dropdown 
menus or encoded categories. By using a dropdown menu for 
\texttt{complaint\_type}, data entry can be standardized, reducing 
inconsistencies caused by manual input errors. Additionally, categorical 
variables with a limited set of possible values can be stored more 
efficiently by assigning unique codes or numerical representations for 
each category. For instance, instead of repeatedly storing the full text 
for every complaint type (e.g., ``Noise - Residential,'' ``Noise - 
Commercial''), a single code could represent each type. This not only 
reduces file size but also improves data processing speed and efficiency 
during analysis. Moreover, standardizing categorical inputs prevents 
the inclusion of misspellings or unrecognized entries, ensuring cleaner 
and more reliable data. This approach is particularly useful in large 
datasets, where frequently repeated categorical variables can 
significantly increase storage requirements. Encoding such fields 
as dropdown options or coded variables can lead to faster query times 
and more consistent data handling.

@@

The Data Dictionary could benefit from an update. We found several 
discrepancies between the Data Dictionary and the actual data, which 
could potentially mislead analytic efforts.

\begin{itemize}
    \item While the Open Data portal lists the data types of various 
    fields, the Data Dictionary does not. Nor does it provide details 
    on specific fields, such as constraints or the domain of legal 
    values. For example, the \texttt{incident\_zip} field is specified 
    as ``text.'' This may not be the most suitable definition; 
    ``categorical,'' with a note that the data contains only numeric 
    characters and is not subjected to arithmetic operations, might 
    be a better description.

    \item Many of the domains of legal values are incomplete or 
    inaccurate. For example, the Data Dictionary indicates that the 
    \texttt{status} field has values of \textit{assigned, canceled, 
    closed}, or \textit{pending}, but we also found values such as 
    \textit{in progress, started}, and \textit{unspecified}. 
    Additionally, no SRs were found with a status of 
    \textit{canceled}. The \texttt{address\_type} field similarly 
    lists values of \textit{address, blockface, intersection, latlong}, 
    and \textit{placename}, but we found additional values such as 
    \textit{bbl} and \textit{unrecognized}, and no \textit{latlong} 
    values were observed. Other fields, like \texttt{facility\_type}, 
    \texttt{vehicle\_type}, \texttt{taxi\_pick\_up\_location}, 
    \texttt{road\_ramp}, and \texttt{city}, exhibit similar inaccuracies 
    in their specified domain values.

@@

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields like 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data.

@@

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government-released datasets. The NYC 311 data revealed several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields. 
As mentioned, updating the Data Dictionary to include descriptions of all fields, 
including the computed fields, e.g. \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
be made to ensure that any changes in data structure or content are 
reflected. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight or noon likely stem from automated processes that 
distort the accurate capture of SR durations. By implementing 
real-time validation tools, governments can prevent such anomalies 
from entering the dataset. Automated systems can be designed to 
flag inconsistencies, incorrect values, or illogical sequences, or 
illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user-friendly formats. Additionally, 
governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous closed dates) are discovered and explaining 
how such issues will be resolved. By ensuring transparency and 
keeping the public informed, governments foster trust and promote 
wider use of datasets for research, civic engagement, and innovation.

@@

% \begin{itemize}
% \item  Undertake a dedicated effort 
%     to investigate these data anomalies, confirm the veracity of the 
%     issues, and collaborate with the appropriate Agency Open Data 
%     Representatives to find resolutions. This effort may be complex 
%     and require software changes. Coordination and mutual approval 
%     of changes will be key, requiring strong support from senior 
%     leadership.

%     \item Standards and protocols should be established, agreed upon, 
%     and enforced. This includes defining the domain of legal values, 
%     ensuring data accuracy, implementing address normalization, 
%     clarifying data usage, and providing detailed data definitions.

%     \item Resolve many potential data integrity issues
%     through enforcement by software. For instance, software should 
%     prevent invalid entries like erroneous zip codes. Similarly, fields 
%     like \texttt{created\_date} and \texttt{closed\_date}, which 
%     measure responsiveness, should have logic checks to prevent 
%     illogical sequences, such as an SR being ``closed'' before 
%     being ``opened.''

%     \item The large spikes in SR creation and closure at midnight and 
%     noon likely point to automated processes from external Agency 
%     integrations. These distortions should be eliminated in favor of 
%     capturing accurate SR durations with precise timestamps.

%     \item As discussed, there are six ``computed'' fields that are not 
%     listed in the Data Dictionary. Some of these fields contain 
%     significant inaccuracies. We recommend updating the Data 
%     Dictionary and, if necessary, masking these fields from public use 
%     until their accuracy is improved.

%     \item Duplicate fields should be reviewed for removal. In some 
%     cases, such as \texttt{borough} and \texttt{park\_borough}, the 
%     duplication is 100\%. In others, the match is 80-90\%. 
%     These near-duplicates should be assessed to determine their 
%     relevance in the 311 SR dataset. Initial analysis suggests that 
%     removing these fields could reduce file size by up to 12\%, with 
%     further reductions possible upon deeper review.

%     \item Fields with very low data presence, such as 
%     \texttt{taxi\_company\_borough}, populated only 0.06\% of the time, 
%     should be evaluated to determine if they are necessary.

%     \item The Data Dictionary, last updated in March 2023, should be 
%     revised based on the issues discussed in Section~\ref{sec:datadictionary}.

%     \item The precision of latitude, longitude, and location fields 
%     should be reduced. We recommend limiting precision to four 
%     decimal places (0.000X), which represents approximately 36.4 feet. 
%     More decimal points strain credibility.

%     \item The 11 address-related fields (\texttt{incident\_address}, 
%     \texttt{street\_name}, \texttt{city}, \texttt{intersection\_street\_1 \& 2}, 
%     \texttt{cross\_street\_1 \& 2}, \texttt{landmark}, \texttt{road\_ramp}, 
%     \texttt{bridge\_highway\_segment}, and \texttt{taxi\_pick\_up\_location}) 
%     should be standardized to USPS format. Software can easily 
%     perform this standardization.

%     \item There are multiple ways missing data is represented, such as 
%     ``UNKNOWN'', ``N/A'', ``NA'', ``na'', and blanks (`` ''). Protocols 
%     should be established to standardize the representation of missing data.
% \end{itemize}

@@

This research further emphasizes the necessity of harmonizing data
across city agencies, particularly in resolving discrepancies in field
names, formats, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in agency structures
or reporting standards can introduce subtle biases. Long-term data
consistency is crucial for longitudinal studies and predictive
modeling, as even minor data shifts can lead to significant deviations
in model outcomes, as noted by \citet{rahm2000data} and
\citet{borgman2012conundrum}.

>>>>

Collectively these six City agencies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the ``big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

\begin{figure}[tbp]
    \centering
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs Created Exactly on the Hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \par\medskip
    \begin{subfigure}{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs Closed Exactly on the Hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    \caption{Comparison of SRs Created and Closed by Hour of the Day}
    \label{fig:stacked}
\end{figure}

@@

\texttt{agency} \& \texttt{agency\_name}: The \texttt{agency} field contains 
abbreviations of City agencies (e.g., NYPD, DOT, DSNY), while 
\texttt{agency\_name} contains the full names. Including both seems 
redundant, especially given the space taken up by full names in the 
CSV file. Additionally, the agency abbreviations are easily recognized.

\texttt{landmark} \& \texttt{street\_name}: Per the Data Dictionary, the
\texttt{landmark} field ``can refer to any noteworthy location, 
including but not limited to, parks, hospitals, airports, sports 
facilities, performance spaces, etc.'  However, most entries 
are actually street names, with a 62\% match to \texttt{street\_name}. 
Even non-matches (excluding blanks) appear to technically be
a match, differing only in spelling or formatting 
(e.g., ``NINTH AVE'' vs. ``9 AVE''). This indicates a possible 
misuse of the \texttt{landmark} field as well as a high level of redundancy.

\texttt{cross\_street\_1/2} \& \texttt{intersection\_street\_1/2}: These 
two sets of street pairs are used to help identify the 
\texttt{incident address}. We found 88\% of these pairs to be 
duplicates. This duplication raises the question of which field-pair 
should be trusted; and if they are different, how should they be treated?
It is possible that these two street pairs originated from different 
agencies that used different variable names to represent the same 
information, but no documentation exists to clarify this 
relationship. These two street-pairs are highly redundant.

@@

Table~\ref{tab:summary1} presents a summary of the matching results 
for \texttt{cross\_street\_1} and \texttt{intersection\_street\_1}. As 
the table shows, 88.15\% of the non-blank entries are exact matches.

@@

By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 39.5\%, which for this large dataset equates to 
a reduction of 1.4 Gb, which is significant. A smaller file size means faster downloads, 
less storage impact, and simpler data analysis efforts.

Below is a list of duplicate and near-duplicate fields.

\begin{itemize}
    \item \textbf{\texttt{agency\_name}}: Since each row contains the \texttt{agency} 
    field, which uses abbreviations that are clear and well understood, we believe the
    \texttt{agency\_name} can be removed with minimal data loss.

    \item \textbf{\texttt{park\_borough}}: This field is a 100\% match with 
    the \texttt{borough} field and can be deleted without data loss.

    \item \textbf{\texttt{location}}: This field is a straight concatenation of 
    the latitude and longitude fields. It can be deleted without data loss.

    \item \textbf{\texttt{cross\_street\_1/2}} and \textbf{\texttt{intersection\_street\_1/2}}: 
    These two stree pairs both exhibit an 88\% match. We recommend deleting the 
    \texttt{intersection\_street\_1/2} fields while acknowledging 
    the concomitant data loss.

@@

Additionally, certain fields may not be useful for broad analysis
due to their sparse data population. This especially applies to the fields below which
are predominantly used by the TLC and are 99\% (or greater) blank. 
\begin{quote}
\textbf{Note:} \textit{Even a field that is 99\% blank still contains 64,000 
non-blank rows in this large (6.4 million-row) dataset.}
\end{quote}

\begin{itemize}
    \item \texttt{taxi\_company\_borough}
    \item \texttt{road\_ramp}
    \item \texttt{vehicle\_type}
    \item \texttt{due\_date}
    \item \texttt{bridge\_highway\_direction}
    \item \texttt{bridge\_highway\_name}
    \item \texttt{bridge\_highway\_segment}
    \item \texttt{taxi\_pick\_up\_location}
\end{itemize}

Further reduction is possible by encoding certain categorical 
variables Using \texttt{complaint\_type} as an example,
instead of repeatedly storing the full tex of each complaint
type, it would use a numeric representation. 
(e.g., ``NOISE - STREET/SIDEWALK'')could be represented by
, a numeric code (e.g. 126). In this datase the example complaint type
occurs over 300,000 times. Such encoding not only reduces file 
size, but also improves processing efficiency. Moreover, 
standardizing categorical inputs prevents the inclusion of 
misspellings or unrecognized/invalid entries, ensuring cleaner 
and more reliable data. This approach is particularly useful in large 
datasets, where frequently repeated text-based categorical variables can 
significantly increase storage requirements.

@@

The Data Dictionary would benefit from an update. We found several 
discrepancies between the Data Dictionary and the actual data, which 
could potentially mislead analytic efforts. Here are some examples:

\begin{itemize}
    \item While the Open Data portal lists the data types of various 
    fields, the Data Dictionary does not. Nor does it provide details 
    on specific fields such as constraints or the domain of legal 
    values. For example, in the Data Dictionary, the \texttt{incident\_zip} 
    field is specified as ``text.'' This may not be the most suitable 
    definition. Perhaps ``categorical,'' with a note indicating that 
    the data must contain only numeric values and is not subjected to 
    arithmetic operations would be a better dictionary description.

    \item Many of the domains of legal values are incomplete or 
    inaccurate. For example, the Data Dictionary indicates that the 
    \texttt{status} field has values of \textit{assigned, canceled, 
    closed}, or \textit{pending}. However, we found additional
    values such as \textit{in progress, started}, and \textit{unspecified}. 
    Additionally, no SRs were found with a status of 
    \textit{canceled}.

    \item The \texttt{address\_type} field similarly 
    lists values of \textit{address, blockface, intersection, latlong}, 
    and \textit{placename}. However, we found additional values such as 
    \textit{bbl} and \textit{unrecognized}. Additionally, no \textit{latlong} 
    values were observed. Other fields, such as \texttt{facility\_type}, 
    \texttt{vehicle\_type}, \texttt{taxi\_pick\_up\_location}, 
    \texttt{road\_ramp}, and \texttt{city}, exhibit similar inaccuracies 
    in their specified domain values.

@@

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields such as 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data. Additionally, a number of fields have 
very defined domains, such as \texttt{incident\_zip}. Such fields
should be subject to data entry validation, in this example
validating against the USPS database at the data entry point.

@@

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government-released datasets. Our study found several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields, and 
inaccurate domain values for categorial fields such as \texttt{status}. 
As mentioned, updating the Data Dictionary to include descriptions of all fields, 
including the computed fields, e.g. \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight (or noon) likely stem from a cron-job type 
processes that distort the accurate capture of SR 
durations. By implementing real-time validation tools, governments 
can prevent such anomalies from entering the dataset in the first 
place. Automated 
systems can be designed to flag inconsistencies, incorrect values, 
or illogical sequences, or illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user-friendly formats. Additionally, 
governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous \texttt{closed\_ date(s)} are discovered, including
an explanation about how such issues will be resolved. By 
ensuring transparency and keeping the public informed, 
governments foster trust and promote wider use of datasets 
for research, civic engagement, and innovation.

@@

@@

This research further emphasizes the necessity of harmonizing data
across city agencies, particularly in resolving discrepancies in field
names, formats, field usage, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in agency structures
or reporting standards can introduce subtle biases. Long-term data
consistency is crucial for longitudinal studies and predictive
modeling, as even minor data shifts can lead to significant deviations
in model outcomes, as noted by \citet{rahm2000data} and
\citet{borgman2012conundrum}.
